---
title: "R for beginners 2"
author: "Huong & Marlene"
output: html_notebook
---

### Agenda:

+ Softwares installation (in case you do not have R or Rstudio installed)
+ About the database
+ Exploration and manipulation
+ Missing values
+ Outliers
+ Some filtering
+ Questions/Discussion


---

### Softwares installation

If you do not have R or Rstudio intalled, use the followig links

- Download and install R: [https://www.r-project.org/]

- Download and install Rstudio: [https://www.rstudio.com/products/rstudio/download/]


---

### About the database

We are going to work with a database from STATBEL, the Belgian statistical office, which describes the population by place of residence, nationality, marital status, age and sexstructure.  

Download the data set from here (txt file recommended):

https://statbel.fgov.be/en/open-data/population-place-residence-nationality-marital-status-age-and-sex-8



Verify your work directory 
```{r}
getwd()
```

if necessary set your work directory to the path where you saved the  database. In my case :
```{r}
input = "C:/Users/marlene/Documents/Rladies/RLadiesWorkshops/rbiginners2"

setwd(input)
```

Load the dataset into R
```{r pop}

pop <- read.table("TF_SOC_POP_STRUCT_2018.txt", header=T, sep="|", stringsAsFactors = F,  encoding = "UTF-8")

```

### Exploration and manipulation
The dataset is quite big, do not call it directly. Let's see the first five rows

```{r}
head(pop, 5)
```

what about the dimension ?

```{r}
dim(pop)
```

and what about the structure of the dataset ?

```{r}
str(pop)
```


#### **Your turn !**

*Exercice 1*:change the name of column 20, from "CD_AGE" to "age"

Hint : use colnames()

```{r}
colnames(pop)
```

```{r}
# Solution
#which(colnames(pop)=="CD_AGE")
colnames(pop)[20] <- "age"
```

### More functions

```{r}
typeof(pop)
```
```{r}
class(pop)
```

```{r}
length(pop)

```

```{r}
pop[20]
```



### Missing Values

We have missing values when no data value is stored for a variable. 
Let's verify is the dataset has missing values.
```{r}
anyNA(pop)
```
How many missing values do we have ?
```{r}
sum(is.na(pop))
```
```{r}
sum(is.na(pop$age))
```


#### **Your turn !**

*Exercice 2*: Where are the missig values ?

Hint : you can use sapply or which functions
```{r}
apply(is.na(pop),2,sum)
```


```{r}
which(is.na(pop),arr.ind=T)
```  

```{r}
pop[58208,7]
```
### Outliers

Outliers are those data points that are really far from the rest. We use the mean as reference.

```{r}
summary(pop[1])
```

Formally outliers are usually defined as  observations that are more than three standard deviations from the mean. 

We can visualize outliers using the boxplot 

```{r}
boxplot(pop[1])
```



#### **Your turn !**

*Exercice 3*: Use boxplot() function to try to find if the dataset has outliers

Hint : verify only colums with int values

```{r}
boxplot(pop[21])
```

Note: We will show later a more clever solution

### Some filtering

Get only females and keep all other columns

```{r}
female_pop = pop[pop$CD_SEX == "F", ]
head(female_pop, 7) 
```

get females and foreigners

```{r}
female_foreign = female_pop[female_pop$CD_NATLTY == "ETR", ]
head(female_foreign,7)
```

get males

```{r}
males_foreign =pop[pop$CD_NATLTY == "ETR" & pop$CD_SEX == "M", ]
head(males_foreign,7)
```

Now let's see some contingency tables. For instance, here we have a two dimensional table which show the number of female and males by regions

```{r}
 table(pop$TX_PROV_DESCR_FR, 
      pop$CD_SEX)
```

a three dimensional table
```{r}
 table(pop$TX_RGN_DESCR_FR, 
      pop$CD_NATLTY,
      pop$CD_SEX)
```

### What we have learned today? 

New R functions:

- class()   retrieves the internal class of an object
- is.na()   indicates which elements are missing (logical)
- anyNA()   basically implements any(is.na(x)).
- which()   returns the position of the elements in a logical vector 
- apply()   it applys a function to margins of an array or matrix
- summary() used to produce result summaries
- boxplot() produce box-and-whisker plot(s) 
- table()   builds a contingency table 

### Thanks for comming !!!
