---
title: "R for beginners"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
  output: html_notebook
---
### Softwares installation ###

- Download and install R: [https://www.r-project.org/]

- Download and install Rstudio: [https://www.rstudio.com/products/rstudio/download/]

### Some exercises in R terminal ###
Adding two numbers together

```{r}
5+5
```
Making a variable

```{r}
a = 5+5  # assigning integer
a
a + 2
a # you are not assigning it so it won't show up

b = 3-6
b

c = a + b 

# h = 5 + g

# 3pi = 3 + 11

# go% = 5 + 6
```
As you can see there are limits on how you can name your variables. Let's make some more variables.


Strings in a list

```{r}
monty = c("eric", "terry", "gilliam")

pie = c("gilliam", "4", "6", "terry")

# not.pie = c("gilliam", 4, "6", "terry", john)
```

### Now let move to Rstudio and play with a data set###

We can see a list of available data sets using *data()* function.

```{r}
data()
```

To see what is in the data set we can use function *View()*.
```{r}
Orange # use View(Orange)
```

Use function *dim()* to see the dimension.

```{r}
dim(Orange)
```
and *str()* to see the structure of the R object
```{r}
str(Orange)
```
More functions to manipulate a data set

```{r}
typeof(Orange$Tree) # 5 level, integer type
typeof(Orange$age) # another word for integer, we will get more into it late
typeof(Orange$circumference) # simply integer
```
See the first 3 rows of this data set, use function *head()* with comma 3
```{r}
head(Orange, 3)
```
See the last 3 rows with function *tail()* 
```{r}
tail(Orange, 3)
```
If you are interested in just looking at 1 variable, you can use dollar sign to select it
```{r}
Orange$age
```
Now we are going to explain row and column, use bracket to select rows and columns
```{r}
Orange[3, 2]
```
Select another cell with row and column
```{r}
Orange[10, 3]
```
Select multiple rows
```{r}
Orange[1:5, ]
```
Selecting multiple rows,
*c()* allows you to make your own selection
```{r}
Orange[c(1:3, 7:15), ]
```
Select 2 variables
```{r}
Orange[c(1:3, 7:15), 2:3]
```
Delete a column
```{r}
Orange[, -1]
```
Delete rows
```{r}
Orange[-(1:3), ]
#Orange[-c(1:3,9:11), ]
```

Let's start assigning, there are two ways: <- or = 
They do the same thing, it depends on your preference.
```{r}
y <- Orange$age[1:10]

x = Orange$age[5:15]
```

You should make a copy of the data set before you make any changes.
So we make a copy of Orange to Orange 1, then we change levels to character.
```{r}
Orange1 = Orange

Orange1$Tree = as.character(Orange1$Tree)
```

Observe that all the integers in the variable Tree have become strings. Now we assign a name to each tree type.

```{r}
Orange1$Tree[1:7] = "oak"
Orange1$Tree[8:14] = "pine"
Orange1$Tree[15:21] = "willow"
Orange1$Tree[22:28] = "cherry"
Orange1$Tree[Orange1$Tree == 5] = "elm"

```
Let's do some filtering. For instance, we are only interested in age more than 500 or between 100 and 1000.
```{r}
tree_age500 = Orange1[Orange1$age > 500,]

tree_age1000 = Orange1[Orange1$age > 1000,]

tree_age5001000 = Orange1[Orange1$age > 500 & Orange1$age < 1000,]
```
Look for help if you need it.
```{r, results="hide"}
help("sort")
?sort
```
Difference between default parameters and changing them
```{r}
tree_age1000$age = sort(tree_age1000$age, decreasing = T)
```
Summary of age of tree
```{r}
summary(Orange1$age)
```
Summary of age of pine
```{r}
summary(Orange1$age[Orange1$Tree == "pine"])
```
Makke data frames
```{r}
small = data.frame(tree = Orange1$Tree,
                   age = Orange1$age)

total.age.circumference = data.frame(circ = rowsum(Orange1$circumference, group = Orange1$Tree), 
                                     age = rowsum(Orange1$age, group = Orange1$Tree))
```
Let's draw a plot.
```{r}
plot(Orange1$age, Orange1$circumference )
```
You can run a linear model. But don't use samples of this size; it is too small - 35 trees! this is bad statistics.
```{r}
lm(formula = Orange1$circumference ~ Orange1$age)
```

### Some useful references ###
Basic R
https://github.com/rladies/meetup-presentations_brussels/blob/master/R_tutorials/introR.pdf

Advanced R
http://adv-r.had.co.nz/

[Base R cheet Sheets](https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf)

A First Course in Statistical Programming with R, W. John Braun and
Duncan J. Murdoch


