---
title: "Writing R package documentation with Roxygen2"
author: "Marlene Marchena"
date: "April 20, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Why to write R packages ?

* Best way to reuse  code and to share code with others
* Reproducible research

### Basic structure of an R package

* A `DESCRIPTION` file, where you store important metadata about your package
* An `R/` directory, where you put .R files
* A `man` directory, where you store .Rd files
* A`NAMESPACE` file, it helps to avoid naming conflicts among packages

### DESCRIPTION file

```
Package: Package name
Type: Package
Title: What the Package Does (Title Case)
Version: 0.1.0
Author: Who wrote it
Maintainer: The package maintainer <yourself@somewhere.net>
Description: More about what it does (maybe more than one line)
    Use four spaces when indenting paragraphs within the Description.
License: What license is it under?
Encoding: UTF-8
LazyData: true
```


### R directory

In this directory you put your R fuctions.

Some examples:

```{r}
hello <- function() {
  print("Hello, world!")
}
```

```{r}
mysum <- function(x,y) x+y
```

```{r}
myproduct<- function(x,y) x*y
```

### man directory

In the man (for manual) directory you need to add the documentation for each function in a special .Rd format (.Rd for R documentation). The .Rd format is based on Latex sintaxis.

Here some examples:

```
\name{hello}
\alias{hello}
\title{Hello, World!}
\usage{
hello()
}
\description{
Prints 'Hello, world!'.
}
\examples{
hello()
}
```

### NAMESPACE file

- R package name system to organize object with identical name . It helps    encapsulate your package and makes it self-contained.

- For instance, both dplyr and plyr packages provides an `arrange()` function. If you load dplyr, then `arrange()` will refer to the dplyr package. But if you load the plyr package first, `arrange()` will refer to the plyr version. <br>

- When using the operator `::` the order in which the packages are loaded won’t matter. In our example you can explicitly refer to specific functions  using:`dplyr::arrange()` or `plyr::arrange()`. 

- Namespaces make your packages self-contained in two ways: the imports and the exports. 
 
- The imports defines how a function in one package finds a function in another

- The exports helps you avoid conflicts with other packages by specifying which functions are available outside of your package  

### NAMESPACE file example

From <http://r-pkgs.had.co.nz/namespace.html#namespace>

```
# Generated by roxygen2 (4.0.2): do not edit by hand
S3method(as.character,expectation)
S3method(compare,character)
export(auto_test)
export(auto_test_package)
export(colourise)
export(context)
exportClasses(ListReporter)
exportClasses(MinimalReporter)
importFrom(methods,setRefClass)
useDynLib(testthat,duplicate_)
useDynLib(testthat,reassign_function)
```


Each namespace directive describes an R object. A directive says if an object is exported from your package to be used by others, or it’s imported from another package to be used locally

### Namespace directives

```
# Export functions (including S3 and S4 generics)
export()
# export all functions that match a pattern
exportPattern()
# Export S4 classes and methods
exportClasses(), exportMethods():
# export S3 methods
S3method()
#import all functions from a package
import()
#import selected functions (including S4 generics)
importFrom()
#import S4 classes and methods
importClassesFrom(), importMethodsFrom()
#import a function from C. This is described in more detail in compiled code
useDynLib()
```

### Why to use Roxygen2?

* Writing .Rd file by hand is painful. Roxygen2 generates it automatically 

* You describe your functions in comments next to their definitions so 
 when you modify your code, you’re reminded to also update your documentation

* Roxygen2 dynamically inspects the objects that it documents


* It manages your NAMESPACE file

* It abstracts over the differences in documenting different types of objects, so you need to learn fewer details.


### Roxygem2 examples

```{r}
#' Documentation title
#' 
#' More description
#' @param x parameter documentation.
#' @param y parameter documentation.
#' @return Description of the return value.
#' @examples
#' Some usage example
#` @export
func <- function(x, y) {
  ...
}
```

```{r}
#' Hello world
#' 
#' Print Hello world
#'
hello <- function() {
  print("Hello, world!")
}
```
### Roxygen tags

Some roxygen2 tags 

```
@title Title of the function
@author author's name
@param declapermet de déclarer les paramètres de la fonction.
@return value to be returned by the function
@references bibliography to help the reader
@note more information about the function
@examples some examples to be runned
@export allows that the function will be available to others.It adds the function to the NAMESPACE

```

### Important links

[https://cran.r-project.org/web/packages/roxygen2/](https://cran.r-project.org/web/packages/roxygen2/)

[https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html](https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html)

[http://kbroman.org/pkg_primer/](http://kbroman.org/pkg_primer/)
